An Efficient Forecasting Approach\\ to Reduce Boundary Effects in Real-Time Time-Frequency Analysis

### 2021-02-22 Paper/theory.tex
polish
final version

\subsection{Signal Model}
\label{sse:model.sine}
We model the meaningful part of the observed signal as a deterministic multicomponent harmonic signal; that is, a sum of sine waves:
\begin{equation}
\bz[n]=\sum_{j=1}^J\Omega_j\cos\left(2\pi f_j \frac{n}{\fs} + \varphi_j\right)\ ,
\label{eq:sum.sine}
\end{equation}
where $J$ denotes the number of components, $\Omega_j>0$ the amplitude of the $j$-th component, $f_j$ its frequency, and $\varphi_j\in[0,2\pi)$ its initial phase.
%
For the sake of simplicity, we make an additional assumption on the frequencies of each component. We assume that for all $j\in\{1,\dots,J\}$:
\begin{equation}
\exists\, p_j,p_j'\in\NN^*: \quad f_j = \dfrac{p_j}{M}\fs = \dfrac{p'_j}{K}\fs\ .
\end{equation}

@@

The forecasting bias and covariance asymptotically depend linearly on the ratio $\frac{1}{K}$. This shows the need to use a sufficiently large dataset to obtain an accurate forecast. Ideally when $K\to\infty$, the forecasting error would behave like the measurement noise $\sigma\bw$, \ie~a zero-mean white noise whose variance is of the order of $\sigma^2$. However, Theorem~\ref{th:error} shows that the estimator may remain asymptotically biased (the first term in the bound of equation~\eqref{eq:mean.error} being independent of $K$). Concerning the covariance of the forecasting error, the expected asymptotic behavior is verified. Indeed, when $K \to\infty$, the variance of the forecasting estimator increases at most linearly with the noise variance $\sigma^2$, since is bounded by $b_0^{(n,n')}\sigma^2$.

@@

The dependencies of the forecasting quality on the subsignals lengths $M$ and the forecasting index $\ell=n-N+1$ are hidden in the expression of the parameters $a^{(n)}_0$, $a^{(n)}_1$,\dots, $b^{(n,n')}_0$,\dots, $c^{(n,n')}_3$. The numerical results, discussed in section~\ref{ssse:res.sine}, allow us to better evaluate these dependencies.

@@

The models of sections~\ref{sse:model.sine} and~\ref{RemarkAHM} consider the meaningful part of the signal to be a deterministic component. These models are purposely adapted to signals showing local line spectra. The physiological signals we are interested in, such as respiratory or cardiac signals, have this characteristic (see section~\ref{sse:physio.sig}). Signals with wider local spectra, such as electroencephalography signals, do not fall into this category, and are more faithfully modeled as random signals. Thus, the theoretical justifications proposed above are no longer applicable to guarantee the forecasting quality of {\sf SigExt}.

\subsection{Performance of the Boundary Effects Reduction}
\label{sse:perf.BoundEffRed}
While we do not provide a generic proof of the boundary effect reduction on {\em any} TF analysis tools, we discuss a particular case of SST. Since SST is designed to analyze signals satisfying the AHM, as is discussed in section~\ref{RemarkAHM}, Theorem~\ref{th:error} ensures that the forecasting error $\bepsilon$ defined in~\eqref{eq:forecasting.error} is controlled and bounded in terms of mean and covariance. Recall Theorem 3 in~\cite{Chen14nonparametric}, which states that when the additive noise is stationary and may be modulated by a smooth function with a small fluctuation, the SST of the observed signal remains close to the SST of the clean signal, throughout the TF plane. 
%
We refer the reader to~\cite{Chen14nonparametric} for a precise quantification of the error made in the TF plane, which depends notably on the covariance of the additive noise and on the speed of variation of the amplitudes and instantaneous frequencies composing the signal. 
%
In our case, while the noise of the historical data is stationary, the forecasting error depends on the historical noise, and hence the overall ``noise'' is nonstationary. However, the dependence only appears in the extended part of the signal. We claim that the proof of Theorem 3 in~\cite{Chen14nonparametric} can be generalized to explain the robustness of SST to the noise, and a systematic proof will be reported in our future theoretical work. We verify experimentally that Algorithm~\ref{alg:boundary} is efficient for a large number of representations in the following section~\ref{se:results}. 
%
Therefore, in our case, this means that the boundary effect is strongly reduced since the impact of the forecasting error does SST is limited. An immediate application is that the instantaneous frequencies can now be well estimated continuously up to the edges of the TF plane, and real-time acquisition of the instantaneous frequency and amplitude information is possible.

>>>>

\subsection{Signal Model}
\label{sse:model.sine}
We model the meaningful part of the observed signal as a deterministic multicomponent harmonic signal; that is, a sum of sine waves:
\begin{equation}
\bz[n]=\sum_{j=1}^J\Omega_j\cos\left(2\pi f_j \frac{n}{\fs} + \varphi_j\right)\ ,
\label{eq:sum.sine}
\end{equation}
where $J$ denotes the number of components, $\Omega_j>0$ the amplitude of the $j$th component, $f_j$ its frequency, and $\varphi_j\in[0,2\pi)$ its initial phase.
%
For the sake of simplicity, we make an additional assumption on the frequencies of each component. We assume that for all $j\in\{1,\dots,J\}$:
\begin{equation}
\exists\, p_j,p_j'\in\NN^*: \quad f_j = \dfrac{p_j}{M}\fs = \dfrac{p'_j}{K}\fs\ .
\end{equation}

@@

The forecasting bias and covariance asymptotically depend linearly on the ratio $\frac{1}{K}$. This shows the need to use a sufficiently large dataset to obtain an accurate forecast. Ideally when $K\to\infty$, the forecasting error would behave like the measurement noise $\sigma\bw$, \ie, a zero-mean white noise whose variance is of the order of $\sigma^2$. However, Theorem~\ref{th:error} shows that the estimator may remain asymptotically biased (the first term in the bound of equation~\eqref{eq:mean.error} being independent of $K$). Concerning the covariance of the forecasting error, the expected asymptotic behavior is verified. Indeed, when $K \to\infty$, the variance of the forecasting estimator increases at most linearly with the noise variance $\sigma^2$, since is bounded by $b_0^{(n,n')}\sigma^2$.

@@

The dependencies of the forecasting quality on the subsignals lengths $M$ and the forecasting index $\ell=n-N+1$ are hidden in the expression of the parameters $a^{(n)}_0$, $a^{(n)}_1$,\dots, $b^{(n,n')}_0$,\dots, $c^{(n,n')}_3$. The numerical results, discussed in Section~\ref{ssse:res.sine}, allow us to better evaluate these dependencies.

@@

The models described in Sections~\ref{sse:model.sine} and~\ref{RemarkAHM} consider the meaningful part of the signal to be a deterministic component. These models are purposely adapted to signals showing local line spectra. The physiological signals we are interested in, such as respiratory or cardiac signals, have this characteristic (see Section~\ref{sse:physio.sig}). Signals with wider local spectra, such as electroencephalography signals, do not fall into this category, and are more faithfully modeled as random signals. Thus, the theoretical justifications proposed above are no longer applicable to guarantee the forecasting quality of {\sf SigExt}.

\subsection{Performance of the Boundary Effects Reduction}
\label{sse:perf.BoundEffRed}
While we do not provide a generic proof of the boundary effect reduction on {\em any} TF analysis tools, we discuss a particular case of SST. Since SST is designed to analyze signals satisfying the AHM, as is discussed in Section~\ref{RemarkAHM}, Theorem~\ref{th:error} ensures that the forecasting error $\bepsilon$ defined in~\eqref{eq:forecasting.error} is controlled and bounded in terms of mean and covariance. Recall Theorem 3 in~\cite{Chen14nonparametric}, which states that when the additive noise is stationary and may be modulated by a smooth function with a small fluctuation, the SST of the observed signal remains close to the SST of the clean signal, throughout the TF plane. 
%
We refer the reader to~\cite{Chen14nonparametric} for a precise quantification of the error made in the TF plane, which depends notably on the covariance of the additive noise and on the speed of variation of the amplitudes and instantaneous frequencies composing the signal. 
%
In our case, while the noise of the historical data is stationary, the forecasting error depends on the historical noise, and hence the overall ``noise'' is nonstationary. However, the dependence only appears in the extended part of the signal. We claim that the proof of Theorem 3 in~\cite{Chen14nonparametric} can be generalized to explain the robustness of SST to the noise, and a systematic proof will be reported in our future theoretical work. We verify experimentally that Algorithm~\ref{alg:boundary} is efficient for a large number of representations in the following Section~\ref{se:results}. 
%
Therefore, in our case, this means that the boundary effect is strongly reduced since the impact of the forecasting error does SST is limited. An immediate application is that the instantaneous frequencies can now be well estimated continuously up to the edges of the TF plane, and real-time acquisition of the instantaneous frequency and amplitude information is possible.
